Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
small edits to paper

Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide a software implementation of our methodology to aid in the application of our techniques to other experimental data sets.
\vspace{1em}

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

@@

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
%
However, pairwise alignments can be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

@@

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
There are some registration errors in the early images, which are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
%
%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

%
The benefits of our approach to image data mining are twofold.
%
First, the algorithm can accomplish the tasks of registration and ordering in a single step.
%
Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis \citep{shlens2005tutorial} versus vector diffusion maps for the data sets presented in this paper).
%
We expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
%
The main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied.
%
For such data sets, manual ordering of the images can be nontrivial, and our algorithms can clearly accelerate uncovering the underlying developmental dynamics.

>>>>

Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide software to aid in the application of our methodology to other experimental data sets.
\vspace{1em}

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

@@

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
%
However, pairwise alignments can also be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

@@

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
%
%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

%
The benefits of our approach to image data mining are twofold.
%
First, the algorithm can accomplish the tasks of registration and ordering in a single step.
%
Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis versus vector diffusion maps for the data sets presented in this paper).
%
We expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
%
The main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied.
%
For such data sets, manual ordering of the images can be nontrivial, and our algorithms can clearly accelerate uncovering the underlying developmental dynamics.
